Community engagement and subgroup meta-knowledge: Some factors in the soul of a community

### 2018-12-31 PaperFinalDraft.Rnw
fix reference
additional reviewer comments

\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\RequirePackage{fix-cm}
\smartqed% flush right qed marks, e.g. at end of proof
\usepackage{graphicx}
\usepackage[round]{natbib}

@@

\begin{abstract} The Knight Foundation collects data to determine what factors impact community satisfaction, local GDP growth, and interest in Knight news publications. For the 2013 Data Expo at the Joint Statistical Meetings, many participants created graphical explorations of these data. This article focuses on the idea of community meta-knowledge, which is essentially majority group empathy or understanding of how minorities experience their community. For example, the survey asks participants to rate their community ``as a place for senior citizens,'' on a 5-point Likert scale. A city where seniors rated their community in the same way as non-seniors is defined as a community with high meta-knowledge about conditions for seniors. Three minority groups were explored: seniors, families with young children, and racial minorities. In most communities, people outside the minority group tended to under-rate their community, compared to those in the minority group. However, there were some exceptions. 
\keywords{2013 Data Exposition, R, ggplot2, Likert scales, meta-knowledge}
\end{abstract}

\section{Introduction}
\label{intro}
Studies have shown that increasing empathy is the best way to improve intergroup relations \citep{SteFin1999}. Therefore, it is of interest to quantify the typical level of empathy in communities across the United States. The Knight Foundation data provides a window to a factor which could be thought of as a proxy for empathy, namely community meta-knowledge. We are defining meta-knowledge as community awareness by those outside a specific subgroup about the conditions for people inside the subgroup. The primary attempt of this article is to answer the question ``Are people outside a specific subgroup aware of the quality of their community for people in that subgroup?" and then, ``do communities with high meta-knowledge (those where people outside the subgroup understood conditions for minorities) have higher community satisfaction rates than those with low meta-knowledge?"

This article is one of several related to the Knight Foundation community data from the 2013 American Statistical Association Data Expo. For more information on the Expo and the data sets, see \citep{Hof2019}. This article is fully reproducble. The paper was written using \LaTeX\, and R~\citep{R2014}. The R package knitr was used to interweve writing and R code~\citep{Xie2014}. If you inspect the source code, available on Github~\citep{McN2019}, you will find that each of the figures in the paper are generated when the document is compiled (``knit''). Likewise, any statistic mentioned in the text is calculated using R and automatically inserted into the text using knitr.  All graphics were generated with the package ggplot2, an R implementation of The Grammar of Graphics~\citep{Wic2009, Wil2005}. The paper was written in the time period between the development of the R package plyr~\citep{Wic2011} and the R package dplyr, which superceded it~\citep{WicFra2015}. Therefore, it uses functions from both packages. One challenge with reproducibility is that as packages develop, they are not always backwards-compatible. To help combat this issue, the code available on Github includes a packrat library. The packrat package allows a user to snapshot the state of their packages in order to freeze them for future reproducibility~\citep{UshMcP2014}. For example, although the current version of dplyr is 0.7.8, this paper uses version 0.4.1. In order to run the code hosted on Github, you will need to ``hydrate'' the library of packages using packrat. The Github repository contains the underlying data, the LaTeX/knitr file which produces the paper (including R code), a packrat archive of the necessary R packages, and LaTeX style files to create the specific journal style.

The remainder of this paper is an exploration of the idea of community meta-knowledge. It begins with discussion of the data in Section \ref{data}, then moves to focus on community satisfaction in Section \ref{communittsatsec}, and community engagement through local behaviors in Section \ref{behaviorsec}. Finally, it discusses meta-knowledge in Section \ref{metasec} and wraps up with conclusions in Section \ref{conclusion}.

@@

While the Gallup Poll analysis is interesting, it does raise the question of multicollinearity, as factors that are correlated with Community Attachment may simply be correlated with one of the variables that was used to compose it, and may not actually have an impact on local GDP growth. 
<<packages, echo=FALSE>>=
require(knitr)
@ 
<<chunkoptions, echo=FALSE, include=FALSE>>=
opts_chunk$set(echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.align='center')
require(dplyr)
require(ggplot2)
require(scales)
require(RColorBrewer)
require(ggmap)
require(xtable)
require(reshape2)
require(forcats)
@
<<readdata>>=
sotc10 <- read.csv("data/sotc10.csv")
sotc09 <- read.csv("data/sotc09.csv")
sotc08 <- read.csv("data/sotc08.csv")
@
<<renaming>>=
renameData <- function(datasetname){
  plyr::rename(datasetname,c("INTRO1" = "gender", "QSB" = "citystate", "QS3"="county", "QS3_02"="county2", "QS3A"="zipcode", "QD1"="age", "QS4"="area_descrip", "QS5"="further", "QS5_2"="further2", "QN1A"="current_life", "QN1B"="future_life", "QCE1"="community_sat","QCE2"="recommend","Q3A"="proud", "Q3B"="perfect", "Q3C"="reputation", "Q4_1"="problem_1", "Q4_2"="problem_2", "Q4_3"="problem_3", "Q4_1_1"="problem_1", "Q4_1_2"="problem_2", "Q4_1_3"="problem_3","Q4_2_1"="problem_4", "Q4_2_2"="problem_5", "Q4_2_3"="problem_6","Q4_3_1"="problem_7", "Q4_3_2"="problem_8", "Q4_3_3"="problem_9","Q5"="rather_live", "Q6"= "five_years_ago", "Q6A" = "five_years_future", "Q7A"= "parks", "Q7B"= "beauty", "Q7C"= "highways", "Q7D"= "affordable_housing", "Q7E" =  "job_opportunites", "Q7F"=  "public_schools", "Q7G" = "colleges", "Q7H"= "nightlife", "Q7I"= "friends", "Q7J" = "raise_kids", "Q7K" = "heathcare", "Q7L" ="leadership", "Q7M" = "care", "Q7N"= "police", "Q7O" = "arts", "Q7P" =  "community_events", "Q8A" = "talented_grads", "Q8B" = "immigrants", "Q8C"= "minorities", "Q8D"= "families_kids", "Q8E" = "gay_lesbian", "Q8F" = "seniors", "Q8G" = "young_adults", "Q9" = "economy", "Q10" = "econ_better", "Q10A" = "trust_government", "Q11" = "employment", "Q12" = "commute", "Q13" = "job_satisfaction", "Q14" = "company_workforce", "Q15" = "enough_income", "Q15AA" = "area_hiring", "Q15AB" = "representative_leaders", "Q16A" = "treated_respectfully", "Q16B" = "well_rested", "Q16C" = "high_stress", "Q16D" = "learned_yest", "Q17" = "stress_source", "Q18" = "night_safety", "Q19" = "crime", "Q20" = "crime_increasing", "Q21"=  "registered_vote", "Q22A" = "volunteer", "Q22B" = "public_meeting", "Q22C"= "voted_local", "Q22D" = "worked_change", "Q22E" = "church", "Q22F" = "festival", "Q22G" = "donated_org", "Q22H" = "donated_individual", "Q22I" = "donated_shelter", "Q22_A" = "impact", "Q23" = "clubs", "Q24" = "friends_nearby", "Q24A" =  "friends_friends", "Q25"= "family_nearby", "Q26"= "talk_neighbors", "QD2" = "years_lived", "QD2A" ="permanent_resident", "QD3" = "work_cat", "QD3A" = "household_adults", "QD4" = "children", "QD5A" = "under6", "QD5B" = "six_12", "QD5C" = "thirteen_17", "QD6" = "married", "QD7"= "education", "QD8" = "own_home", "QD9"= "income", "QD10" = "hispanic", "QD111" = "race", "QD112" = "race2", "QD113" = "race3", "QD12" = "white_hispanic", "QD13" = "phone_numbers", "QD13A" = "cell_only", "QD13B" = "have_cell", "QD13C" = "cell_calls"))
}

@@

<<calculatingInGroupFeelings>>=
FeelingsC <- function(dataset){
  Feelings <- dataset %>% 
    group_by(citystate, InGroup) %>% 
    summarise(percents= sum(Freq))
  Feelings <- data.frame(Feelings)
  newDF <- data.frame(InGroup=NA, OutGroup=NA, citystate=levels(Feelings$citystate))
  newDF$ratio <- (Feelings[Feelings$InGroup=="FALSE" | Feelings$InGroup=="No",]$percents) - (Feelings[Feelings$InGroup=="TRUE" | Feelings$InGroup=="Yes",]$percents)
  newDF$InGroup <- Feelings[Feelings$InGroup=="TRUE" | Feelings$InGroup=="Yes",]$percents
  newDF$OutGroup <- Feelings[Feelings$InGroup=="FALSE" | Feelings$InGroup=="No",]$percents
  return(newDF)
}
@

@@

\subsection{Missing Data}
\label{missingdata}
While the Gallup reports claim the telephone surveys only took 15 minutes, the number of variables collected and the wide range of response scales seems to indicate a much larger time commitment. This raises the question of whether everyone who began the survey completed it. And, as mentioned in Section \ref{maprates}, the 2010 data contained many more observations than previous years. The explanation for this difference is missing data, presumably related to surveys that were not fully completed.

@@

A followup to this question is whether all communities performed these actions at similar rates, or if there were local variations in behavior. Figure \ref{fig:YN2010plot} shows the percentage-point difference from the overall rate across all 26 communities and 10 behaviors in 2010. For example, in 2010, approximately \Sexpr{round(trial1[["pos"]][trial1[["pos"]]$Year=="2010" & trial1[["pos"]]$Question=="registered_vote",]$Freq, digits=2)*100}\% of people arcross all communities were registered to vote. In comparison, in Aberdeen, SD, \Sexpr{round(dfcitiesYes[dfcitiesYes$citystate=="Aberdeen, SD" & dfcitiesYes$Question=="registered_vote",]$Freq, digits=2)*100}\% of people reported being registered to vote, so Figure \ref{fig:YN2010plot} shows a percentage-point difference of +\Sexpr{round(dfcitiesYes[dfcitiesYes$citystate=="Aberdeen, SD" & dfcitiesYes$Question=="registered_vote",]$Freq-trial1[["pos"]][trial1[["pos"]]$Year=="2010" & trial1[["pos"]]$Question=="registered_vote",]$Freq, digits=2)*100}. Of course, all the communities rates are included in the calculation of the average, so it makes sense that some communities are above the overall rate and some are below. More study is required to determine if the visual differences between communities represent true differences or just random variation, but there are certainly communities that stand out from the rest.

For example, people in Boulder, CO, Long Beach, CA, and San Jose, CA participated in chuch events at a much lower rate the the overall, while people in Georgia (both Macon and Milledgeville) were more likely to.

It is also interesting to note the communities that performed above or below national rates on most questions. For example, Biloxi, MS shows that respondents engage in almost all of the behaviors asked about at higher-than average rates, with the exception of being registered to vote and performing local volunteer work. On the other hand, Palm Beach, FL shows lower-than-average rates for almost all behavior, except for giving money or food to an individual. Long Beach, CA shows a similar pattern. However, since this array of plots represents 26 facets of the same plot, it is highly possible that these visual trends occurred by chance. Humans have a tendency to see patterns in noise, even when no true pattern exists. This tendency is known as apophenia or pareidolia, and can be combatted by the use of visual inference techniques~\citep{BujCoo200}. However, that was outside the scope of this Data Expo entry.

@@

\begin{eqnarray}
\label{MK}
MK = \sum R^{+}_{\mbox{ In subgroup}} - \sum R^{+}_{\mbox{ Out of subgroup}}
\end{eqnarray}

@@

Using this definition, communities where Whites highly over-rate their community as a place for minorities will have a negative $MK_S$ score, while communities where Whites under-rate their community will have a positive $MK_S$ score. Communities where the ratings by both groups are roughly equal have a $MK_S$ score of zero.

For a visual explanation of this rating system, see Figure \ref{fig:anotherlookplot}. Points above the $y=x$ line will have a negative $MK$ score, those below the line will have a positive score, and those on the line will be zero.

@@

For the most part, however, Whites over-rated their communities as a place for racial and ethnic minorities, compared to Non-whites. Grand Forks, ND, was a particularly bad offender-- Whites over-rated it as a place for minorities, and minorities themselves rated it as one of the worst communities in 2009.

@@

For another view of the relationship between ratings (as seen in Figure \ref{fig:allMinorities}) see Figure \ref{fig:anotherlookplot}, which shows the relationship between total positive responses by Whites versus total positive responses by Non-whites to the question, ``How is your community as a place for racial and ethnic minorities?'' Figure \ref{fig:anotherlookplot} makes it clear that there are communities that are under- and over-rated by Whites, and those that are closer to the 1-1 (or y=x) line.

@@

However, even if a community is on the 1-1 line in Figure \ref{fig:anotherlookplot}, it may fall below the mean rating for communities overall. Gary, IN is a good example. It has almost the same overall positive rating from Whites and Non-whites (around \Sexpr{round(newDF09[newDF09$citystate=="Gary, IN",]$nonwhite, digits=2)})\% positive ratings), but falls below the mean ratings of communities overall. So, there is agreement that Gary is a worse place than average for minorities. Of course, looking at Figure \ref{fig:allMinorities}, we can see that using the overall positive rating is obscuring some aspects of the overall distribution. Although the positive end of the scale matches up quite nicely, the distribution of negative responses by non-whites is much heavier tailed. As with all simple measures, focusing just on the total positive percentage loses information.

@@

al <- ggplot(newDF09) + 
  geom_polygon(data=df_poly, aes(x, y), fill="grey", alpha=0.6) +
  geom_vline(aes(xintercept=vert), color="grey") + 
  geom_hline(aes(yintercept=horiz), color="grey") + 
  geom_abline(intercept=0, slope=1) + 
  geom_point(aes(y=white, x=nonwhite)) + 
  geom_text(
    data = newDF09[newDF09$citystate %in% c("Duluth, MN", "Gary, IN", "Grand Forks, ND", "State College, PA"),],
    aes(y=white, x=nonwhite, label = citystate), 
    vjust=c(-0.6, -0.4, 1.3, -0.6), 
    hjust=0, 
    size=4
    ) + 
  xlab("Non-whites rating") + 
  ylab("Whites rating") + 
  xlim(0.4, 1) + 
  ylim(0.4, 1) 
al
@

@@

<<diffscor>>=
diffscor <- cor(masterData[masterData$year != 2010,]$communitysat, abs(masterData[masterData$year != 2010,]$racediffs))
@

@@

The correlation suggests a positive relationship between the variables, but it's not what we would expect. Communities that had the highest absolute difference in scoring of the question ``how is your community as a place for minorities?'' between Whites and Non-whites (and therefore the highest $|MK_R|$ score) had the highest community satisfaction scores. Communities that had low meta-knowledge score (like Gary, IN) have low community satisfaction scores. This seems counter-intuitive, as we would expect communities that had more self awareness to be more satisfied, but that's not what we see in the plot. It is clear that there is more to community satisfaction than just $|MK_R|$.

@@

The yearly ratings distributions are shown in Figure \ref{fig:seniorOverallPlot}. It appears that non-seniors typically underestimate how good a place is for seniors. People 62 and older rated their community as a better place for seniors than did people under 62, over all three years.

@@

baseplot <- ggplot() + 
  facet_wrap(~citystate, ncol=3) + 
  geom_bar(mapping = aes(InGroup, Freq, fill = Response, order = Response), data = seniorGroup09$neg, stat = "identity") + 
  geom_bar(mapping = aes(InGroup, Freq, fill = Response, order = Response), data = seniorGroup09$pos, stat = "identity") + 
  scale_fill_manual(
    breaks=c("Very bad", "2", "3", "4", "Very good"), 
    values=colorsA, 
    name="Response"
    ) + 
  coord_flip() + 
  scale_y_continuous(
    breaks=seq(from=-0.25, to=0.75, by=0.25), 
    labels=percent
    ) + 
  scale_x_discrete(labels=c("Non-seniors", "Seniors")) + 
  xlab("") + 
  ylab("") + 
  theme(
    legend.text=element_text(size=14), 
    legend.title=element_text(size=16), 
    axis.text.y=element_text(size=16), 
    title=element_text(size=16), 
    axis.text.x=element_text(size=12),
    strip.text=element_text(size=14)) +
  geom_text(data=totalSeniors09, aes(label=paste0("(",totalN, ")"), x=InGroup, y=1.2), size=3, hjust=1) +
  expand_limits(y=1)

@@

<<anotherlookseniors, out.width="0.76\\linewidth", fig.cap=paste0("Relationship between positive responses to the question, \`\`How is your community as a place for seniors?\" comparing ratings of non-seniors and seniors. Each community is represented, and the plot uses 2009 data. The darker grey region corresponds to positive meta-knowledge scores. The black line shows y=x, for comparison. Grey lines at x=", round(mean(totPos$Seniors), digits=2), " and y=", round(mean(totPos$Nonseniors), digits=2), " show the mean ratings by each group.") >>=
sr <- ggplot(totPos) + 
  geom_polygon(data=df_poly, aes(x, y), fill="grey", alpha=0.6) +
  geom_vline(aes(xintercept=vert), color="grey") + 
  geom_hline(aes(yintercept=horiz), color="grey") + 
  geom_abline(intercept=0, slope=1) + 
  geom_point(aes(y=Nonseniors, x=Seniors)) + 
  geom_text(
    data = totPos[totPos$citystate %in% c("Wichita, KS", "Gary, IN", "Bradenton, FL"),], 
    aes(y=Nonseniors, x=Seniors, label = citystate), 
    vjust=c(-0.4,-0.4,1), 
    hjust=c(1,0,0), 
    size=4
    ) + 
  xlab("Senior rating") + 
  ylab("Non-senior rating") +
  ylim(0.5, 1) +
  xlim(0.5, 1)

>>>>

\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\RequirePackage{fix-cm}
\smartqed% flush right qed marks, e.g. at end of proof
\usepackage{graphicx}
\usepackage[round]{natbib}
\usepackage{amsmath}

@@

\begin{abstract} The Knight Foundation collects data to determine what factors impact community satisfaction, local GDP growth, and interest in Knight news publications. For the 2013 Data Expo at the Joint Statistical Meetings, participants created graphical explorations of these data. This article focuses on the idea of community meta-knowledge, which is essentially majority group empathy or understanding of how minorities experience their community. For example, the survey asks participants to rate their community ``as a place for senior citizens,'' on a 5-point Likert scale. A city where seniors rated their community in the same way as non-seniors is defined as a community with high meta-knowledge about conditions for seniors. Three minority groups were explored: seniors, families with young children, and racial minorities. In most communities, people outside the minority group tended to under-rate their community, compared to those in the minority group. However, meta knowledge about racial minorities stood out as an exception. 
\keywords{2013 Data Exposition, R, ggplot2, Likert scales, meta-knowledge}
\end{abstract}

\section{Introduction}
\label{intro}
Studies have shown that increasing empathy is the best way to improve intergroup relations \citep{SteFin1999}. Therefore, it is of interest to quantify the typical level of empathy in communities across the United States. The Knight Foundation data provides a window into something which could be thought of as a proxy for empathy, namely community meta-knowledge. We define meta-knowledge as community awareness by those outside a specific subgroup about the conditions for people inside the subgroup. The primary attempt of this article is to answer the question ``Are people outside a specific subgroup aware of the quality of their community for people in that subgroup?" and then, ``do communities with high meta-knowledge (those where people outside the subgroup understood conditions for minorities) have higher community satisfaction rates than those with low meta-knowledge?"

This article is one of several related to the Knight Foundation community data from the 2013 American Statistical Association Data Expo. For more information on the Expo and the data sets, see \cite{Hof2019}.

This article is fully reproducible. The paper was written using \LaTeX\, and R~\citep{R2014}. The R package knitr was used to interweave writing and R code~\citep{Xie2014}. If you inspect the source code, available on Github~\citep{McN2019}, you will find that each of the figures in the paper are generated when the document is compiled (``knit''). Likewise, any statistic mentioned in the text is calculated using R and automatically inserted into the text using knitr.  All graphics were generated with the package ggplot2, an R implementation of The Grammar of Graphics~\citep{Wic2009, Wil2005}. The paper was written in the time period between the development of the R package plyr~\citep{Wic2011} and the R package dplyr, which superseded it~\citep{WicFra2015}. Therefore, it uses functions from both packages.

One challenge with reproducibility is that as packages develop, they are not always backwards-compatible. To help combat this issue, the code available on Github includes a packrat library. The packrat package allows a user to snapshot the state of their R packages in order to freeze them for future reproducibility~\citep{UshMcP2014}. For example, although the current version of dplyr is 0.7.8, this paper uses version 0.4.1. In order to run the code hosted on Github, you will need to ``hydrate'' the library of packages using packrat. The Github repository contains the underlying data, the \LaTeX/knitr file which produces the paper (including R code), a packrat archive of the necessary R packages, and \LaTeX\, style files to create the specific journal style.

The remainder of this paper is an exploration of the idea of community meta-knowledge. It begins with a discussion of the data in Section \ref{data}, then moves to focus on community satisfaction in Section \ref{communittsatsec}, and community engagement through local behaviors in Section \ref{behaviorsec}. Finally, it discusses meta-knowledge in Section \ref{metasec} and wraps up with conclusions in Section \ref{conclusion}.

@@

While the Gallup Poll analysis is interesting, it does raise the question of multicollinearity, as factors that are correlated with Community Attachment may simply be correlated with one of the variables that were used to compose it, and may not actually have an impact on local GDP growth. 
<<packages, echo=FALSE>>=
require(knitr)
@ 
<<chunkoptions, echo=FALSE, include=FALSE>>=
opts_chunk$set(echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.align='center')
require(dplyr)
require(ggplot2)
require(scales)
require(RColorBrewer)
require(ggmap)
require(xtable)
require(reshape2)
require(forcats)
@
<<readdata>>=
sotc10 <- read.csv("data/sotc10.csv")
sotc09 <- read.csv("data/sotc09.csv")
sotc08 <- read.csv("data/sotc08.csv")
@
<<renaming>>=
renameData <- function(datasetname){
  plyr::rename(datasetname,c("INTRO1" = "gender", "QSB" = "citystate", "QS3"="county", "QS3_02"="county2", "QS3A"="zipcode", "QD1"="age", "QS4"="area_descrip", "QS5"="further", "QS5_2"="further2", "QN1A"="current_life", "QN1B"="future_life", "QCE1"="community_sat","QCE2"="recommend","Q3A"="proud", "Q3B"="perfect", "Q3C"="reputation", "Q4_1"="problem_1", "Q4_2"="problem_2", "Q4_3"="problem_3", "Q4_1_1"="problem_1", "Q4_1_2"="problem_2", "Q4_1_3"="problem_3","Q4_2_1"="problem_4", "Q4_2_2"="problem_5", "Q4_2_3"="problem_6","Q4_3_1"="problem_7", "Q4_3_2"="problem_8", "Q4_3_3"="problem_9","Q5"="rather_live", "Q6"= "five_years_ago", "Q6A" = "five_years_future", "Q7A"= "parks", "Q7B"= "beauty", "Q7C"= "highways", "Q7D"= "affordable_housing", "Q7E" =  "job_opportunites", "Q7F"=  "public_schools", "Q7G" = "colleges", "Q7H"= "nightlife", "Q7I"= "friends", "Q7J" = "raise_kids", "Q7K" = "heathcare", "Q7L" ="leadership", "Q7M" = "care", "Q7N"= "police", "Q7O" = "arts", "Q7P" =  "community_events", "Q8A" = "talented_grads", "Q8B" = "immigrants", "Q8C"= "minorities", "Q8D"= "families_kids", "Q8E" = "gay_lesbian", "Q8F" = "seniors", "Q8G" = "young_adults", "Q9" = "economy", "Q10" = "econ_better", "Q10A" = "trust_government", "Q11" = "employment", "Q12" = "commute", "Q13" = "job_satisfaction", "Q14" = "company_workforce", "Q15" = "enough_income", "Q15AA" = "area_hiring", "Q15AB" = "representative_leaders", "Q16A" = "treated_respectfully", "Q16B" = "well_rested", "Q16C" = "high_stress", "Q16D" = "learned_yest", "Q17" = "stress_source", "Q18" = "night_safety", "Q19" = "crime", "Q20" = "crime_increasing", "Q21"=  "registered_vote", "Q22A" = "volunteer", "Q22B" = "public_meeting", "Q22C"= "voted_local", "Q22D" = "worked_change", "Q22E" = "church", "Q22F" = "festival", "Q22G" = "donated_org", "Q22H" = "donated_individual", "Q22I" = "donated_shelter", "Q22_A" = "impact", "Q23" = "clubs", "Q24" = "friends_nearby", "Q24A" =  "friends_friends", "Q25"= "family_nearby", "Q26"= "talk_neighbors", "QD2" = "years_lived", "QD2A" ="permanent_resident", "QD3" = "work_cat", "QD3A" = "household_adults", "QD4" = "children", "QD5A" = "under6", "QD5B" = "six_12", "QD5C" = "thirteen_17", "QD6" = "married", "QD7"= "education", "QD8" = "own_home", "QD9"= "income", "QD10" = "hispanic", "QD111" = "race", "QD112" = "race2", "QD113" = "race3", "QD12" = "white_hispanic", "QD13" = "phone_numbers", "QD13A" = "cell_only", "QD13B" = "have_cell", "QD13C" = "cell_calls"))
}

@@

<<calculatingInGroupFeelings>>=
FeelingsC <- function(dataset){
  Feelings <- dataset %>% 
    group_by(citystate, InGroup) %>% 
    summarise(percents= sum(Freq))
  Feelings <- data.frame(Feelings)
  newDF <- data.frame(InGroup=NA, OutGroup=NA, citystate=levels(Feelings$citystate))
  # just changed, because subtraction was in the wrong direction...
  newDF$ratio <- (Feelings[Feelings$InGroup=="TRUE" | Feelings$InGroup=="No",]$percents) - (Feelings[Feelings$InGroup=="FALSE" | Feelings$InGroup=="Yes",]$percents)
  newDF$InGroup <- Feelings[Feelings$InGroup=="TRUE" | Feelings$InGroup=="Yes",]$percents
  newDF$OutGroup <- Feelings[Feelings$InGroup=="FALSE" | Feelings$InGroup=="No",]$percents
  return(newDF)
}
@

@@

\subsection{Missing Data}
\label{missingdata}
While the Gallup reports claim the telephone surveys only took 15 minutes, the number of variables collected and the wide range of response scales seem to indicate a much larger time commitment. This raises the question of whether everyone who began the survey completed it. And, as mentioned in Section \ref{maprates}, the 2010 data contained many more observations than previous years. The explanation for this difference is missing data, presumably related to surveys that were not fully completed.

@@

A followup to this question is whether all communities performed these actions at similar rates, or if there were local variations in behavior. Figure \ref{fig:YN2010plot} shows the percentage-point difference from the overall rate across all 26 communities and 10 behaviors in 2010. For example, in 2010, approximately \Sexpr{round(trial1[["pos"]][trial1[["pos"]]$Year=="2010" & trial1[["pos"]]$Question=="registered_vote",]$Freq, digits=2)*100}\% of people across all communities were registered to vote. In comparison, in Aberdeen, SD, \Sexpr{round(dfcitiesYes[dfcitiesYes$citystate=="Aberdeen, SD" & dfcitiesYes$Question=="registered_vote",]$Freq, digits=2)*100}\% of people reported being registered to vote, so Figure \ref{fig:YN2010plot} shows a percentage-point difference of +\Sexpr{round(dfcitiesYes[dfcitiesYes$citystate=="Aberdeen, SD" & dfcitiesYes$Question=="registered_vote",]$Freq-trial1[["pos"]][trial1[["pos"]]$Year=="2010" & trial1[["pos"]]$Question=="registered_vote",]$Freq, digits=2)*100}. Of course, all the communities rates are included in the calculation of the average, so it makes sense that some communities are above the overall rate and some are below. More study is required to determine if the visual differences between communities represent true differences or just random variation, but there are certainly communities that stand out from the rest.

For example, people in Boulder, CO, Long Beach, CA, and San Jose, CA participated in church events at a much lower rate the the overall, while people in Georgia (both Macon and Milledgeville) were more likely to participate in church events.

It is also interesting to note the communities that performed above or below national rates on most questions. For example, Biloxi, MS shows that respondents engage in almost all of the behaviors asked about at higher-than average rates, with the exception of being registered to vote and performing local volunteer work. On the other hand, Palm Beach, FL shows lower-than-average rates for almost all behavior, except for giving money or food to an individual. Long Beach, CA shows a similar pattern. However, since this array of plots represents 26 facets of the same plot, it is highly possible that these visual trends occurred by chance. Humans have a tendency to see patterns in noise, even when no true pattern exists. This tendency is known as apophenia or pareidolia, and can be combated by the use of visual inference techniques~\citep{WicCoo2010}. However, that was outside the scope of this Data Expo entry.

@@

\begin{eqnarray}
\label{MK}
MK = \frac{1}{\sum R^{+}_{\mbox{ In subgroup}} - \sum R^{+}_{\mbox{ Out of subgroup}}}
\end{eqnarray}

This equation means that communities with similar scores from people inside and outside the subgroup will receive high absolute $MK$ values, while those with very different scores will receive low absolute $MK$ values. A positive $MK$ score means that people outside the subgroup are over-rating the community, while a negative $MK$ score means they are under-rating it.

@@

Using this definition, communities where Whites highly over-rate their community as a place for minorities will have a negative $MK_R$ score, while communities where Whites under-rate their community will have a positive $MK_R$ score. Communities where the ratings by both groups are roughly equal will have a large $MK_R$ score, whether positive or negative. Generally, we are more interested in the magnitude of the score than the sign, but it is sometimes interesting to consider which group is under- or over-rating.

For a visual explanation of this rating system, see Figure \ref{fig:anotherlookplot}. Points close to the $y=x$ line have a large $MK$ score, those below the line will have smaller positive score, and those above the line have smaller negative scores.

@@

For the most part, however, Whites over-rated their communities as a place for racial and ethnic minorities, compared to Non-whites. Grand Forks, ND, was a particularly bad offender-- Whites over-rated it as a place for minorities (negative $MK$ score), and minorities themselves rated it as one of the worst communities in 2009.

@@

For another view of the relationship between ratings (as seen in Figure \ref{fig:allMinorities}) see Figure \ref{fig:anotherlookplot}, which shows the relationship between total positive responses by Whites ($\sum R^+_{\text{out of subgroup}}$) versus total positive responses by Non-whites ($\sum R^+_{\text{in subgroup}}$) to the question, ``How is your community as a place for racial and ethnic minorities?'' Figure \ref{fig:anotherlookplot} makes it clear that while there are some communities that are under-rated by Whites (positive $MK$ score), the majority of communities are over-rated by Whites as a place for minorities (negative $MK$ score). Very few communities are close to the 1-1 (or y=x) line (large $|MK|$ score).

@@

However, even if a community is on the 1-1 line in Figure \ref{fig:anotherlookplot}, it may fall below the mean rating for communities overall. Gary, IN is a good example. It has almost the same overall positive rating from Whites and Non-whites (around \Sexpr{round(newDF09[newDF09$citystate=="Gary, IN",]$nonwhite, digits=2)})\% positive ratings), so it has an $MK$ score of \Sexpr{round(1/newDF09[newDF09$citystate=="Gary, IN",]$racediffs)} but it falls below the mean ratings of communities by both Whites and Non-whites. So, while there is high meta-knowledge in Gary, it is agreement that Gary is a worse place than average for minorities. Of course, looking at Figure \ref{fig:allMinorities}, we can see that using the overall positive rating ($\sum R^+$) to compute $MK$ is obscuring some aspects of the overall distribution. Although the positive end of the scale matches up quite nicely, the distribution of negative responses by non-whites is much heavier tailed. As with all simple measures, focusing just on the total positive percentage loses information.

@@

al <- ggplot(newDF09) + 
  geom_polygon(data=df_poly, aes(x, y), fill="grey", alpha=0.6) +
  geom_vline(aes(xintercept=vert), color="grey") + 
  geom_hline(aes(yintercept=horiz), color="grey") + 
  geom_abline(intercept=0, slope=1) + 
  geom_point(aes(y=white, x=nonwhite)) + 
  geom_text(
    data = newDF09[newDF09$citystate %in% c("Duluth, MN", "Gary, IN", "Grand Forks, ND", "State College, PA"),],
    aes(y=white, x=nonwhite, label = paste(citystate, "\nMK =", round(1/racediffs))), 
    vjust=c(-0.4, 1.1, 1.3, -0.4), 
    hjust=0, 
    size=4
    ) + 
  xlab("Non-whites rating") + 
  ylab("Whites rating") + 
  xlim(0.4, 1) + 
  ylim(0.4, 1) 
al
@

@@

<<diffscor>>=
diffscor <- cor(masterData[masterData$year != 2010,]$communitysat, abs(1/masterData[masterData$year != 2010,]$racediffs))
@

@@

The correlation suggests a weak negative relationship between the variables, which is not what we would expect. Communities that had lowest meta-knowledge scores had the highest community satisfaction scores, on average. Communities that had high meta-knowledge have lower community satisfaction scores. This seems counter-intuitive, as we would expect communities that had more self awareness to be more satisfied, but that's not what we see. It is clear that there is more to community satisfaction than just $|MK_R|$.

@@

The yearly ratings distributions are shown in Figure \ref{fig:seniorOverallPlot}. It appears that non-seniors typically underestimate how good a place is for seniors (positive $MK$ score). People 62 and older rated their community as a better place for seniors than did people under 62, over all three years.

@@

baseplot <- ggplot() + 
  facet_wrap(~citystate, ncol=3) + 
  geom_bar(mapping = aes(InGroup, Freq, fill = Response, order = Response), data = seniorGroup09$neg, stat = "identity") + 
  geom_bar(mapping = aes(InGroup, Freq, fill = Response, order = Response), data = seniorGroup09$pos, stat = "identity") + 
  scale_fill_manual(
    breaks=c("Very bad", "2", "3", "4", "Very good"), 
    values=colorsA, 
    name="Response"
    ) + 
  coord_flip() + 
  scale_y_continuous(
    breaks=seq(from=-0.25, to=0.75, by=0.25), 
    labels=percent
    ) + 
  scale_x_discrete(labels=c("Non-seniors", "Seniors")) + 
  xlab("") + 
  ylab("") + 
  theme(
    legend.text=element_text(size=14), 
    legend.title=element_text(size=16), 
    axis.text.y=element_text(size=16), 
    title=element_text(size=16), 
    axis.text.x=element_text(size=11),
    strip.text=element_text(size=14)) +
  geom_text(data=totalSeniors09, aes(label=paste0("(",totalN, ")"), x=InGroup, y=1.2), size=3, hjust=1) +
  expand_limits(y=1)

@@

<<anotherlookseniors, out.width="0.76\\linewidth", fig.cap=paste0("Relationship between positive responses to the question, \`\`How is your community as a place for seniors?\" comparing ratings of non-seniors and seniors. Each community is represented, and the plot uses 2009 data. The darker grey region corresponds to positive meta-knowledge scores. The black line shows y=x, for comparison. Grey lines at x=", round(mean(totPos$Seniors), digits=2), " and y=", round(mean(totPos$Nonseniors), digits=2), " show the mean ratings by each group.") >>=
sr <- ggplot(totPos) + 
  geom_polygon(data=df_poly, aes(x, y), fill="grey", alpha=0.6) +
  geom_vline(aes(xintercept=vert), color="grey") + 
  geom_hline(aes(yintercept=horiz), color="grey") + 
  geom_abline(intercept=0, slope=1) + 
  geom_point(aes(y=Nonseniors, x=Seniors)) + 
  geom_text(
    data = totPos[totPos$citystate %in% c("Wichita, KS", "Gary, IN", "Bradenton, FL"),], 
    aes(y=Nonseniors, x=Seniors, label = paste(citystate, "\nMK =", round(1/seniordiffs))
        ), 
    vjust=c(-0.4,-0.4,1), 
    hjust=c(1,0,0), 
    size=4
    ) + 
  xlab("Senior rating") + 
  ylab("Non-senior rating") +
  ylim(0.5, 1) +
  xlim(0.5, 1)
